---
title: MLCA
author: Hyevin Shin
format: html
---

```{r setup}
## ---------------------------
##' [libraries]
## ---------------------------

library(tidyverse)
library(readxl)
library(tidycensus)
library(multilevLCA)

rm(list = ls())
```


```{r data}
## ---------------------------
##' [IPEDS/Carnegie Classification Data]
## ---------------------------

# load IPEDS data and merge with the Carnegie Classification

ipeds24 <- read_csv("../data/sources/hd2024.csv")
ipeds_cost <- read_csv("../data/sources/cost1_2024.csv")
ipeds_schoolsize <- read_csv("../data/sources/ef2024a.csv") #enrollment Fall 2024
carnegie <- read_excel("../data/sources/2025-Public-Data-File.xlsx", sheet = 2)

ipeds_schoolsize <- ipeds_schoolsize |> 
    filter(EFALEVEL == 1) |> #Level of student (1 == all students total)
    select(UNITID, EFTOTLT) |> #Grand total
    mutate(size = case_when(
      EFTOTLT <= 3000 ~ 0,  #small
      EFTOTLT <= 10000 ~ 1,  #medium
      EFTOTLT > 10000 ~ 2, #large
      TRUE ~ NA_real_))

carnegie <- carnegie |> 
  select(unitid, hsi)

data <- ipeds24 |>  
  left_join(ipeds_cost, by = "UNITID") |> 
  left_join(ipeds_schoolsize, by = "UNITID") |> 
  left_join(carnegie, by = c("UNITID" = "unitid"))



# select variables and create a dataset for analysis
x <- data |> 
    rename_all(tolower) |> #converts all column names in a data frame to lowercase
    select(unitid, instnm, city, countycd, countynm, cbsa, cbsatype, csa, 
           locale, zip, stabbr, longitud, latitude, 
           control, iclevel, instcat, deggrant, size,
           carnegieic, carnegiesaec, carnegiealf,
           hbcu, hsi, tribal, room)


# assign to "data"
data <- x


```


```{r cleaning}
## ---------------------------
##' [Data Cleaning]
## ---------------------------

x <- data |> 
  mutate(
         public = case_when(
            control == 1 ~ 1,                             # 0: private
            control == 2 ~ 0,                             # 1: public 
            TRUE ~ NA_real_),   
         
         twoyear = case_when(
            iclevel == 1 ~ 0,                             # 0: 4 or more
            iclevel == 2 ~ 1,                             # 1: at least 2 & less than 4
            TRUE ~ NA_real_),     
         
         msi = case_when(                                 # 0: non of them
            hbcu == 1 ~ 1,                                # 1: HBCU
            hsi == 1 ~ 2,                                 # 2: HSI
            tribal == 1 ~ 3,                              # 3: TCU
            hbcu == 2 & hsi == 0 & tribal == 2 ~ 0,
            is.na(hbcu) | is.na(hsi) | is.na(tribal) ~ NA_real_,
            TRUE ~ NA_real_),
         
         rurality = case_when(
            locale %in% c(31,32,33,41,42,43) ~ 0,         # 0: Town + Rural
            locale %in% c(21,22,23) ~ 1,                  # 1: Suburb
            locale %in% c(11,12,13) ~ 2,                  # 2: City
            TRUE ~ NA_real_),           

         housing = case_when(
           room == 1 ~ 1,                                 # 0: no housing
           room == 2 ~ 0,                                 # 1: offer housing
           TRUE ~NA_real_),    
         
         asc = case_when(
           carnegiesaec == 1 ~ 0,                         # 0: Lower Access, Lower Earnings
           carnegiesaec == 3 ~ 1,                         # 1: Lower Access, Medium Earnings
           carnegiesaec == 5 ~ 2,                         # 2: Lower Access, Higher Earnings
           carnegiesaec == 2 ~ 3,                         # 3: Higher Access, Lower Earnings 
           carnegiesaec == 4 ~ 4,                         # 4: Higher Access, Medium Earnings
           carnegiesaec == 6 ~ 5,                         # 5: Higher Access, Higher Earnings
           TRUE ~ NA_real_),
#Q. Would it make more sense to reorder this variable so that higher access is coded as a lower level?
                  
         awarelev = case_when(
           carnegiealf == 1:3 ~ 0,                         # 0: associate - baccalaureate
           carnegiealf == 4 ~ 1,                           # 1: undergrad / grad-Master's
           carnegiealf == 5 ~ 2,                           # 2: undergrad / grad-Doctorate 
           TRUE ~ NA_real_),
         
         cbsa = ifelse(cbsa == -2, NA, cbsa), 
         cbsa = as.character(cbsa)
       ) #|> 
    
#    select(-c(control, iclevel, hbcu, hsi, tribal, locale, room, carnegiesaec, carnegiealf))
# Q. Would it be okay to drop all the original variables after recoding them?          
# I’m wondering if it might be better to comment this out and keep the original variables instead of dropping them.

data <- x

```


```{r census/ACS data}
## ---------------------------
##' [Census/ACS Data]
## ---------------------------

census <- get_acs(geography = "cbsa",
                  year = 2024,
                  variables = c("B01003_001E", # Total population
                                "B19013_001E", # Median Household Income in the Past 12 Months
                                "DP03_0119PE", # % below poverty
                                "DP02_0068PE", # % bach or higher
                                "B03002_003E", # total white (alone, not hispanic)
                                "B03002_004E", # total black (alone, not hispanic)
                                "B03002_012E", # total hispanic
                                "B03002_006E", # total asian (alone, not hispanic)
                                "B03002_001E"  # total by ethnicity
                  ),
                  output = "wide") 


census <- census |>
  mutate(perc_white = B03002_003E/B03002_001E*100,
         perc_black = B03002_004E/B03002_001E*100,
         perc_hisp = B03002_012E/B03002_001E*100,
         perc_asian = B03002_006E/B03002_001E*100,
         tot_pop = B01003_001E/100000) |>
  rename(perc_ba = DP02_0068PE,
         med_income = B19013_001E,
         perc_pov = DP03_0119PE) |> 
  select(tot_pop, perc_ba,
         med_income, perc_pov,
         perc_white, perc_black, perc_hisp, perc_asian,
         GEOID) |> 
  mutate(GEOID = as.character(GEOID))

data <- data |>
  left_join(census, by = c("cbsa" = "GEOID"))

```


```{r Create Sample}
level_1_vars <- c("public", "twoyear", "size", "msi", "rurality", "housing", "asc", "carnegiealf", "selectivity")
# we need to decide which variable will be "selectivity"

level_2_vars <- c()

filter(!is.na(control), control != 3) |>              # 3: private for-profit
filter(!is.na(iclevel), iclevel != 3) |>              # 3: less than 2 years
filter(!is.na(carnegiealf), carnegiealf != 6) |>      # 6: graduate-focused
```

Sample criteria
-2-year and 4-year colleges✅
-public or private, not-for-profit✅
-CARNEGIEALF == 1-5✅

Sample notes: 
-remove primarily graduate-degree granting schools 

#Level 1 variables to consider
##indicator as well as covariates
-Carnegie classification (carnegieic,)

The proportion of an CBSA's area that is rural 
-cultural attractions

Drop observations that are not located in CBSAs✅




```{r MCLA}
## ---------------------------
##' [MLCA Analysis Data]
## ---------------------------

data_mlca0 <- data |>
  filter(!is.na(cbsa),
         !is.na(public),
         !is.na(twoyear),
         !is.na(size),
         !is.na(msi),
         !is.na(rurality),
         !is.na(housing),
         !is.na(asc),
         !is.na(awarelev)
         )


glimpse(data_mlca0)
summary(data_mlca0)

data_mlca0 |> 
  count(cbsa, name = "n_institutions") |> 
  summary()
# number of institutions within a CBSA




## ---------------------------
##' [Find the best # of Classes]
## ---------------------------

Y <- c("public", "twoyear", "size", "msi", "rurality", "housing", "asc", "awardlev")    
# indicators
# does "awardlev" make sense as an indicator? or should it be a covariate?

set.seed(1234)

msas0 <- multiLCA(
  data = data_mlca0,
  Y = Y,
  iT = 5:9,               # lower-level classes
#  id_high = "cbsa",
#  iM = 1:4               # higher-level classes
)

# the best number of level1 classes is 7
msas0                     # compare model fits

#when including level2 equations -> Error: sort_index(): detected NaN



## ---------------------------
##' [..... stepwise approach for dealing with the error]
## ---------------------------

set.seed(1234)

msas00 <- multiLCA(
  data = data_mlca0,
  Y = Y,
  iT = 7,
  id_high = "cbsa",
  iM = 1:4
)

# when # of level1 classes is 7, the best number of level2 classes is 3
msas00                     # compare model fits



# CHECK AGAIN

set.seed(1234)

msas000 <- multiLCA(
  data = data_mlca0,
  Y = Y,
  iT = 6:8,
  id_high = "cbsa",
  iM = 3
)

#Error: sort_index(): detected NaN
# Note:
# When tested separately with iM fixed at 3,
# the model converged for iT = 7,
# but did not converge for iT = 6 or iT = 8.
# -> Among the nearby alternatives, the 7-by-3 solution appears to be the only stable candidate

msas000                     # compare model fits



## ---------------------------
##' [Final model]
## ---------------------------

set.seed(1234)

msas1 <- multiLCA(
  data = data_mlca0,
  Y = Y,
  iT = 7,                                  
  id_high = "cbsa",
  iM = 3,
  extout = TRUE #additional outputs
  )             #e.g., classification error, modal class assignment, posterior probabilities

msas1
summary(msas1)
plot(msas1)



## ---------------------------
##' [MLCA including covariates with the best class model]
## ---------------------------

Zh <- c("tot_pop", "perc_ba", "med_income", "perc_pov",
        "perc_white", "perc_black", "perc_hisp", "perc_asian")

data_mlca1 <- data_mlca0 |>
  filter(!is.na(tot_pop),
         !is.na(perc_ba),
         !is.na(med_income),
         !is.na(perc_pov),
         !is.na(perc_white),
         !is.na(perc_black),
         !is.na(perc_hisp),
         !is.na(perc_asian)) |> 
  as.data.frame()


set.seed(1234)

msas2 <- multiLCA(
  data = data_mlca1,
  Y = Y,
  iT = 7,
  id_high = "cbsa",
  iM = 3,
  Zh = Zh,
  extout = TRUE
)

#Error: Zh specified, Z not



# temporary solution
# Q. do we need level-1 covariates? if not, the model doesn't work :(

msas2 <- multiLCA(
  data = data_mlca1,
  Y = Y,
  iT = 7,
  id_high = "cbsa",
  iM = 3,
  Z = Z,             # what variables should be our covariates?
  Zh = Zh,
  extout = TRUE
)


msas2
summary(msas2)
plot(msas2)

```

